---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

```

# insurancerating

The goal of insurancerating is to give analytic techniques that can be used in insurance rating. In particular, the methods can be used to construct tariff classes in P&C insurance. 

## Installation

You can install insurancerating from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("MHaringa/insurancerating")
```

## Example

This is a basic example which shows you how to construct tariff classes for the variable *age_policyholder* in the MTPL dataseet. 


```{r example, eval = TRUE, fig.path = "man/figures/"}
library(insurancerating)
library(ggplot2)

x <- construct_tariff_classes(data = MTPL, nclaims = nclaims, x = age_policyholder, exposure = exposure)

# Show tariff classes
autoplot(x)

```


