---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/"
)
```

# insurancerating

<!-- badges: start -->
[![CRAN Status](https://www.r-pkg.org/badges/version/insurancerating)](https://cran.r-project.org/package=insurancerating)
[![Downloads](https://cranlogs.r-pkg.org/badges/insurancerating?color=blue)](https://cran.rstudio.com/package=insurancerating)
<!-- badges: end -->

The goal of insurancerating is to give analytic techniques that can be used in insurance rating. It provides a data driven strategy for the construction of 
tariff classes in P&C insurance. The goal is to bin the continuous factors such that categorical risk factors result which capture the effect of the covariate on the response in an accurate way, while being easy to use in a generalized linear model (GLM).

It also adds functionality showing additional lines for the reference categories in the levels of the coefficients in the output of a generalized linear regression analysis. In addition it implements a procedure determining the level of a factor with the largest exposure, 
and thereafter changing the base level of the factor to this level. 

## Installation

Install insurancerating from CRAN:

```{r, eval = FALSE}
install.packages("insurancerating")
```

Or the development version from GitHub:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("MHaringa/insurancerating")
```

## Example
This is a basic example which shows the techniques provided in insurancerating. 

The first part shows how to fit a GAM for the variable *age_policyholder* in the MTPL dataset: 

```{r example, eval = TRUE, message = FALSE, warning = FALSE}
library(insurancerating)
library(ggplot2)
library(dplyr)

# Claim frequency 
age_policyholder_frequency <- fit_gam(data = MTPL, 
                                      nclaims = nclaims, 
                                      x = age_policyholder, 
                                      exposure = exposure)

# Claim severity 
age_policyholder_severity <- fit_gam(data = MTPL, 
                                     nclaims = nclaims, 
                                     x = age_policyholder, 
                                     exposure = exposure, 
                                     amount = amount, 
                                     model = "severity")
```

Create plot:

```{r plotgam, eval = TRUE}

autoplot(age_policyholder_frequency, show_observations = TRUE)

```

Determine classes for the claim frequency (the points show the ratio between the observed number of claims and exposure for each age): 

```{r figfreq, eval = TRUE}

clusters_freq <- construct_tariff_classes(age_policyholder_frequency)
clusters_sev <- construct_tariff_classes(age_policyholder_severity)

autoplot(clusters_freq, show_observations = TRUE)
```

The figure shows that younger policyholders have a higher risk profile. The fitted GAM is lower than might be expected from the observed claim frequency for policyholders of age 19. This is because there are very few young policyholders of age 19 present in the portfolio. 

Show classes for the claim severity: 

```{r figsev, eval = TRUE}

age_policyholder_severity %>%
  construct_tariff_classes() %>%
  autoplot(., show_observations = TRUE, remove_outliers = 100000)

```

The second part adds the constructed tariff classes for the variable *age_policyholder* to the dataset, and sets the base level of the factor *age_policyholder* to the level with the largest exposure. In this example for claim frequency the class for ages (39,50], which contains the largest exposure. 

```{r example2, eval = TRUE, message = FALSE, warning = FALSE}


dat <- MTPL %>%
  mutate(age_policyholder_freq_cat = clusters_freq$tariff_classes) %>%
  mutate(age_policyholder_sev_cat = clusters_sev$tariff_classes) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_if(is.factor, list(~biggest_reference(., exposure)))

glimpse(dat)


```

The last part is to fit a *generalized linear model*. The function rating_factors prints the output including the reference group.

```{r example3, eval = TRUE}
model <- glm(nclaims ~ age_policyholder_freq_cat, offset = log(exposure), family = "poisson", data = dat)
rating_factors(model)
```


