---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

`insurancerating` provides functions for common actuarial pricing tasks in R.

A common GLM-based pricing exercise often combines several tasks:

1.	portfolio analysis
2.	model estimation
3.	interpretation of fitted coefficients
4.	refinement of tariff structure

This vignette presents one possible GLM-based analysis and illustrates how the
functions can be combined:

-	analyse risk factors with `factor_analysis()`
-	estimate pricing models with `glm()`
-	interpret coefficients with `rating_table()`
-	assess model stability with `model_performance()` and `bootstrap_performance()`

The focus is on the transition from observed portfolio experience to fitted
effects and a tariff structure that can be reviewed actuarially.


## Data

We use the example dataset `MTPL2`, which contains a motor portfolio with:

-	number of claims (`nclaims`),
-	exposure (`exposure`),
-	premium (`premium`),
-	claim amounts (`amount`),
-	several rating factors
	
```{r, warning = FALSE, message = FALSE}

library(insurancerating)
library(dplyr)

head(MTPL2)

```


## Step 1 — Portfolio analysis

### Factor analysis

A pricing analysis commonly starts with a descriptive review of the portfolio.

Before fitting a model, it is useful to assess:

-	how experience differs across factor levels
-	whether observed differences are supported by sufficient exposure and claim volume
-	whether the observed pattern is plausible

This is done with `factor_analysis()`.


### Basic factor analysis

We start by analysing a single risk factor.

```{r}

fa <- factor_analysis(
  MTPL,
  risk_factors = "zip",
  claim_count = "nclaims",
  exposure = "exposure",
  claim_amount = "amount"
)

fa

```

The output provides commonly used portfolio metrics such as:

-	frequency = claims / exposure
-	average severity = loss / claims
-	risk premium = loss / exposure
-	loss ratio = loss / premium
-	average premium = premium / exposure


### Visualising factor behaviour

```{r}

autoplot(fa, metrics = c("exposure", "frequency", "risk_premium"))

```

This provides a direct view of:

-	the distribution of exposure
-	the variation in claim frequency
-	the variation in risk premium

These are descriptive, univariate results. They show the observed experience
and volume by level, but do not control for correlations with other risk
factors.
	

## Step 2 — Continuous variables

### Why continuous variables are treated separately

Continuous variables can be modelled directly or translated into grouped tariff
variables, depending on the model and implementation environment. This example
uses the following sequence:

1.	analysed as continuous variables
2.	translated into tariff segments
3.	used in a GLM as categorical rating factors

Grouping makes the resulting tariff effect discrete and directly
implementable, but introduces a segmentation choice that should be reviewed.

### Analysing the shape with a GAM

```{r}

age_freq <- risk_factor_gam(
  data = MTPL,
  risk_factor = "age_policyholder",
  claim_count = "nclaims",
  exposure = "exposure"
)

autoplot(age_freq, show_observations = TRUE)

```

This step is used to inspect:

-	non-linear patterns
-	local volatility
-	areas with low exposure
-	plausible breakpoints for tariff segments

### Deriving tariff segments

```{r}

age_segments <- derive_tariff_segments(age_freq)
autoplot(age_segments)
summary(age_segments)

```

This derives candidate segment boundaries from the fitted continuous effect.
Exposure and claim volume have already informed the GAM estimate. They are
reported by `summary()` for review, but are not applied again when the
fitted curve is divided into intervals.
The evolutionary tree approximates the GAM curve rather than the individual
claim outcomes. Each distinct risk-factor value on the fitted curve has equal
influence on this approximation; exposure is not applied as a second weight.

The resulting segments should be reviewed against exposure, observed experience,
stability and practical tariff requirements before they are used in a model.
	
	
### Adding tariff segments to the data

```{r}

dat <- MTPL |>
  add_tariff_segments(age_segments, name = "age_cat") |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(across(where(is.factor), ~ set_reference_level(., exposure)))

```

`set_reference_level()` sets the reference level to the level with the highest
exposure. This changes the coefficient parameterisation, not the fitted values.
A high-exposure level is often a useful baseline because its relativity is
supported by a substantial part of the portfolio.


## Step 3 — Model estimation

### Why GLMs are used

Generalized linear models are widely used in insurance pricing. They can:

-	accommodate non-normal response distributions
-	produce interpretable multiplicative effects
-	can be translated into tariff relativities

A common decomposition is:

- frequency --> Poisson GLM
-	severity --> Gamma GLM

### Frequency model

```{r}

mod_freq <- glm(
  nclaims ~ age_cat,
  offset = log(exposure),
  family = poisson(),
  data = dat
)

```

### Severity model

```{r}

mod_sev <- glm(
  amount ~ age_cat,
  weights = nclaims,
  family = Gamma(link = "log"),
  data = dat |> filter(amount > 0)
)

```

Frequency and severity are modelled separately because they capture different
aspects of the loss process.

### Constructing a premium proxy

```{r}

premium_df <- dat |>
  add_prediction(mod_freq, mod_sev) |>
  mutate(premium = pred_nclaims_mod_freq * pred_amount_mod_sev)

head(premium_df)

```

This produces an expected-loss proxy from the fitted frequency and severity
components. Its precise unit depends on the exposure treatment in the frequency
prediction and should be checked before it is used as a model response.


## Step 4 — Premium model

### Fitting a premium model

```{r}

burn_unrestricted <- glm(
  premium ~ age_cat + zip,
  weights = exposure,
  family = Gamma(link = "log"),
  data = premium_df
)

```

This model combines the rating factors into one fitted risk-premium structure.

It can be used to inspect the combined effect of the frequency and severity
components. Commercial loadings and other premium components are outside this
technical risk-premium model.


## Step 5 — Interpreting coefficients

### Rating table

```{r}

rt <- rating_table(burn_unrestricted)
rt

```

`rating_table()` expresses fitted coefficients in terms of the original factor
levels, including the reference level.

This output is commonly used to inspect tariff relativities.
	

### Visualising coefficients

```{r}

rating_table(burn_unrestricted) |>
  autoplot()

```

This plot can be used to assess:

-	the relative size of coefficients
-	the structure across levels
-	the exposure behind each level
-	whether additional refinement may be needed

At this stage, the relevant questions are:

-	are coefficients sufficiently stable?
-	do they follow the expected pattern?
-	are some levels driven by limited exposure?
	
	

## Step 6 — Model evaluation

### Model performance

```{r}

model_performance(mod_freq)

```

This reports AIC, BIC and response-scale RMSE. These measures are most
meaningful when models use the same response, records, weights and offsets.


### Bootstrap performance

```{r}

bp <- bootstrap_performance(
  mod_freq,
  dat,
  n_resamples = 50,
  sample_fraction = 0.8,
  sampling = "bootstrap",
  show_progress = FALSE
)
autoplot(bp)

```

This refits the model on repeated bootstrap samples and evaluates RMSE on
out-of-bag records. The resulting distribution describes sensitivity to the
sampled portfolio records; it is not a prediction interval for future claims.

A single fit statistic is usually not sufficient. In pricing practice, it is
also relevant to assess whether the model behaves consistently under small data
perturbations.


## Step 7 — From model to tariff

At this point, the example has produced:

-	portfolio-level insight
-	fitted pricing models
-	interpretable factor relativities
-	basic performance diagnostics

In many cases, a further step is required before the model output can be used as
a tariff.

Typical reasons include:

-	irregular coefficient patterns
-	monotonicity requirements
-	externally imposed restrictions
-	expert-driven adjustments

This can be handled with the refinement tools described in [Refinement building blocks](refinement-workflow.html).

## Summary

A possible sequence in `insurancerating` is:

```{r, eval = FALSE}

factor_analysis()             # analyse portfolio behaviour
risk_factor_gam()             # analyse continuous variables
derive_tariff_segments()      # derive tariff segments
glm()                         # estimate pricing models
rating_table()                # interpret fitted coefficients
bootstrap_performance()       # assess stability
prepare_refinement()          # refine tariff structure if needed

```

The sequence distinguishes observed experience, fitted model effects, candidate
tariff segmentation and model diagnostics. The final modelling choices remain
dependent on the portfolio and pricing objective.


## Next steps

The following vignette covers the refinement step in more detail:

- [Refinement building blocks](refinement-workflow.html)

For the conceptual background to exposure, risk premium, and tariff design, see:

- [Pricing workflow building blocks](pricing-workflow-building-blocks.html)
