---
title: "Working with large insurance portfolios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with large insurance portfolios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

database_packages <- c("DBI", "dbplyr", "dplyr", "duckdb")
duckdb_available <- all(vapply(
  database_packages,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))
```

Insurance portfolio extracts can contain millions of policy-period records.
The practical limit is not imposed by R as a statistical language, but by the
memory needed to hold the data and temporary objects created during a
calculation. A database can perform the first reduction when the row-level
portfolio does not fit comfortably in memory.

This vignette compares two workflows:

1. use `rating_grid()` and `merge_date_ranges()` on data already held in R;
2. use `rating_grid_db()` and `merge_date_ranges_db()` to reduce data in a
   database and import only the result.

```{r}
library(insurancerating)
```

## Memory rather than a fixed row limit

There is no generally valid maximum number of rows for R. Required memory
depends on the number and type of columns. Ten million rows containing a few
integer or factor columns are materially smaller than ten million rows with
many character fields. Reading a file and subsequently grouping it may also
require more memory than the final object because source and intermediate
objects coexist temporarily.

The following ranges are practical planning guidance rather than hard limits:

| Portfolio size | Typical approach |
|---|---|
| Up to about 1 million rows | Usually straightforward in R |
| About 1 to 10 million rows | Often feasible for a reasonably narrow table with sufficient memory and `data.table`-based operations |
| Above about 10 million rows | Estimate memory before importing; database reduction is often preferable |
| About 50 million wide portfolio rows | Usually reduce in DuckDB or the source database before collecting into R |

A sample of the source file gives a more useful estimate than its row count:

```{r, eval = FALSE}
bytes_per_row <- as.numeric(object.size(portfolio_sample)) /
  nrow(portfolio_sample)

estimated_object_gb <- bytes_per_row * expected_rows / 1024^3
```

Allow additional working memory for reading, copying, grouping and modelling.
If the estimated object already occupies a substantial part of available RAM,
perform the initial reduction in a database.

## Local portfolio reduction

`rating_grid()` combines records with the same observed rating-factor values.
The following example first rounds reconstruction value to units of EUR 1,000.
This may be appropriate when individual euro values are not relevant to the
tariff structure.

```{r}
set.seed(2026)
local_portfolio <- data.frame(
  policy_id = seq_len(100000),
  sector = sample(c("Industry", "Retail", "Services"), 100000, TRUE),
  region = sample(c("North", "South", "West"), 100000, TRUE),
  reconstruction_value = sample(seq(100000, 2000000, by = 500), 100000, TRUE),
  earned_exposure = runif(100000, 0.25, 1),
  earned_premium = runif(100000, 100, 2500)
)

local_portfolio$reconstruction_value_1000 <-
  round(local_portfolio$reconstruction_value / 1000) * 1000

local_grid <- rating_grid(
  local_portfolio,
  group_by = c("sector", "region", "reconstruction_value_1000"),
  exposure = "earned_exposure",
  aggregate_cols = "earned_premium"
)

data.frame(
  stage = c("Portfolio", "Rating grid"),
  rows = c(nrow(local_portfolio), nrow(local_grid))
)
```

The result contains one row per observed combination and preserves total
exposure and premium. The policy-level table is still required when individual
policy predictions or policy-level diagnostics are needed.

`merge_date_ranges()` performs a different reduction. It combines connected
coverage periods within the same policy or risk:

```{r}
periods <- data.frame(
  policy_id = c("P001", "P001", "P002"),
  coverage = c("Fire", "Fire", "Fire"),
  period_start = as.Date(c("2025-01-01", "2025-07-01", "2025-01-01")),
  period_end = as.Date(c("2025-06-30", "2025-12-31", "2025-12-31")),
  earned_exposure = c(0.5, 0.5, 1)
)

merge_date_ranges(
  periods,
  period_start = "period_start",
  period_end = "period_end",
  group_by = c("policy_id", "coverage"),
  aggregate_cols = "earned_exposure"
)
```

## Lazy reduction in DuckDB

The database functions accept a lazy table created with `dplyr::tbl()`. They
return another lazy table. Calling the function therefore constructs SQL but
does not import the source portfolio.

The next example is executed when the suggested database packages are
available. It writes the generated portfolio to a temporary, file-backed
DuckDB database. The grouping remains lazy: `rating_grid_db()` constructs the
query, and only the reduced grid is read back into R with `collect()`.

```{r duckdb-file-workflow, eval = duckdb_available}
library(DBI)
library(dbplyr)
library(dplyr)
library(duckdb)

database_path <- tempfile(fileext = ".duckdb")
con <- dbConnect(duckdb(), dbdir = database_path)

dbWriteTable(
  con,
  "portfolio",
  local_portfolio,
  overwrite = TRUE
)

portfolio_db <- tbl(con, "portfolio")

grid_db <- portfolio_db |>
  mutate(
    reconstruction_value_1000 =
      round(reconstruction_value / 1000) * 1000
  ) |>
  rating_grid_db(
    group_by = c("sector", "region", "reconstruction_value_1000"),
    exposure = "earned_exposure",
    aggregate_cols = "earned_premium"
  )

# Inspect the SQL without collecting the row-level portfolio.
sql_render(grid_db)

database_row_counts <- data.frame(
  source_rows = portfolio_db |>
    summarise(n = n()) |>
    collect() |>
    pull(n),
  reduced_rows = grid_db |>
    summarise(n = n()) |>
    collect() |>
    pull(n)
)

database_grid <- collect(grid_db)
database_row_counts
head(database_grid)

dbDisconnect(con, shutdown = TRUE)
unlink(database_path)
unlink(paste0(database_path, ".wal"))
```

This example starts from an R object only to make the workflow reproducible
inside the vignette. In production, the row-level portfolio will commonly
already reside in a database or be read by DuckDB directly from Parquet files.
In that case the large source table never needs to be materialised in R.

DuckDB can also query files without first loading them into an R object. For
example, a Parquet extract can be exposed as a database view:

```{r, eval = FALSE}
library(DBI)
library(dbplyr)
library(dplyr)
library(duckdb)

con <- dbConnect(duckdb())

dbExecute(con, "
  CREATE VIEW portfolio AS
  SELECT * FROM read_parquet('portfolio/*.parquet')
")
portfolio_db <- tbl(con, "portfolio")
```

The following DuckDB example generates ten million rows inside the database.
The records are never materialised as an R data frame. The code is not executed
during the vignette build because creating ten million rows on every package
check would be disproportionate, but the block is complete and can be run as
shown.

```{r, eval = FALSE}
large_database_path <- "large_portfolio.duckdb"
con <- dbConnect(duckdb(), dbdir = large_database_path)

dbExecute(con, "
  CREATE TABLE portfolio_10m AS
  SELECT
    i AS policy_id,
    'Sector ' || CAST(i % 20 AS VARCHAR) AS sector,
    'Region ' || CAST(FLOOR(i / 20) % 10 AS VARCHAR) AS region,
    100000 + (FLOOR(i / 200) % 100) * 1000 AS reconstruction_value,
    0.5 + (i % 50) / 100.0 AS earned_exposure,
    100 + (i % 2000) AS earned_premium
  FROM range(10000000) AS portfolio(i)
")

portfolio_10m <- tbl(con, "portfolio_10m")

grid_10m_db <- portfolio_10m |>
  mutate(
    reconstruction_value_1000 =
      round(reconstruction_value / 1000) * 1000
  ) |>
  rating_grid_db(
    group_by = c("sector", "region", "reconstruction_value_1000"),
    exposure = "earned_exposure",
    aggregate_cols = "earned_premium"
  )

sql_render(grid_10m_db)

row_comparison_10m <- data.frame(
  source_rows = portfolio_10m |> summarise(n = n()) |> pull(n),
  reduced_rows = grid_10m_db |> summarise(n = n()) |> pull(n)
)
row_comparison_10m$reduction <-
  1 - row_comparison_10m$reduced_rows / row_comparison_10m$source_rows

row_comparison_10m
grid_10m <- collect(grid_10m_db)
```

In this constructed portfolio, the three grouping variables define at most
20,000 combinations. Ten million source rows are therefore reduced to no more
than 20,000 rating-grid rows before `collect()` is called: a reduction of
99.8%. The exact reduction in a real portfolio depends on the number of
observed combinations.

The same pattern applies to 50 million rows. Only the range used to generate the
database table changes:

```{r, eval = FALSE}
dbExecute(con, "
  CREATE TABLE portfolio_50m AS
  SELECT
    i AS policy_id,
    'Sector ' || CAST(i % 20 AS VARCHAR) AS sector,
    'Region ' || CAST(FLOOR(i / 20) % 10 AS VARCHAR) AS region,
    100000 + (FLOOR(i / 200) % 100) * 1000 AS reconstruction_value,
    0.5 + (i % 50) / 100.0 AS earned_exposure,
    100 + (i % 2000) AS earned_premium
  FROM range(50000000) AS portfolio(i)
")

portfolio_50m <- tbl(con, "portfolio_50m")

grid_50m_db <- rating_grid_db(
  portfolio_50m,
  group_by = c("sector", "region", "reconstruction_value"),
  exposure = "earned_exposure",
  aggregate_cols = "earned_premium"
)

row_comparison_50m <- data.frame(
  source_rows = portfolio_50m |> summarise(n = n()) |> pull(n),
  reduced_rows = grid_50m_db |> summarise(n = n()) |> pull(n)
)

grid_50m <- collect(grid_50m_db)
```

The 50 million source records again reduce to no more than 20,000 combinations
in this example, a reduction of 99.96%. The large table remains in DuckDB; only
the compact grid enters R. A file-backed DuckDB database can be used when the
database itself should persist between sessions.

## Reducing date ranges in DuckDB

`merge_date_ranges_db()` applies the temporal gaps-and-islands calculation in
DuckDB. It returns consolidated periods as a lazy query:

```{r, eval = FALSE}
dbExecute(con, "
  CREATE TABLE portfolio_periods AS
  SELECT * FROM (VALUES
    ('P001', 'Fire', 'Industry', DATE '2025-01-01', DATE '2025-06-30', 0.5, 600.0),
    ('P001', 'Fire', 'Industry', DATE '2025-07-01', DATE '2025-12-31', 0.5, 650.0),
    ('P002', 'Fire', 'Retail',   DATE '2025-01-01', DATE '2025-12-31', 1.0, 900.0)
  ) AS periods(
    policy_id,
    coverage,
    sector,
    period_start,
    period_end,
    earned_exposure,
    earned_premium
  )
")

periods_db <- tbl(con, "portfolio_periods")

merged_periods_db <- merge_date_ranges_db(
  periods_db,
  period_start = "period_start",
  period_end = "period_end",
  group_by = c("policy_id", "coverage", "sector"),
  aggregate_cols = c("earned_exposure", "earned_premium"),
  merge_gap_days = 1
)

merged_periods <- collect(merged_periods_db)
dbDisconnect(con, shutdown = TRUE)
unlink(large_database_path)
unlink(paste0(large_database_path, ".wal"))
```

This database variant is restricted to DuckDB because date arithmetic and
window-function details differ between database systems. `rating_grid_db()` is
based on standard grouped SQL and can be used with other `dbplyr` backends.

## Which reduction comes first?

The two reductions answer different questions and do not always need to be
combined.

- Use `rating_grid_db()` directly when the source periods are already suitable
  for aggregation and the objective is a table of observed rating combinations.
- Use `merge_date_ranges_db()` when renewals, endorsements or administrative
  splits first need to be consolidated into meaningful coverage periods.
- When both are required, merge periods first and construct the rating grid
  second. A rating grid no longer contains the row-level interval structure
  needed to merge policy periods.

The grouping columns used for temporal merging must retain every attribute that
should remain distinct. For example, include `sector` when a policy changes
sector during its history. Aggregate premium or exposure only when the source
amounts are additive; overlapping records that describe the same covered days
should be resolved before summing.

For a large workflow it can be useful to materialise an intermediate reduction
inside DuckDB with `dplyr::compute()`. This avoids repeating an expensive merge
while still keeping the intermediate table outside R. Call `collect()` only
after checking that the reduced row count and columns fit the intended R
analysis.
