---
title: "Model validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Model validation is part of the standard pricing workflow.

After model estimation and coefficient interpretation, it is necessary to assess
whether the model performs adequately and behaves in a stable and interpretable
way.

In practice, model validation typically considers several dimensions:

-	comparative model performance
-	coefficient structure
-	predictive stability
-	distributional diagnostics
-	portfolio-level behaviour

`insurancerating` provides tools for each of these steps.

The purpose of validation is not only to assess statistical fit, but also to
determine whether a model is suitable for use in a pricing context.


## Example setup

The examples below use a simple frequency modelling setup based on `MTPL2`.

```{r}

library(insurancerating)
library(dplyr)

df <- MTPL2 |>
  mutate(across(c(area), as.factor)) |>
  mutate(across(c(area), ~ biggest_reference(., exposure)))

mod1 <- glm(
  nclaims ~ area,
  offset = log(exposure),
  family = poisson(),
  data = df
)

mod2 <- glm(
  nclaims ~ area + premium,
  offset = log(exposure),
  family = poisson(),
  data = df
)

```


## Step 1 — Comparative model performance

A first validation step is to compare alternative model specifications.

```{r}

model_performance(mod1, mod2)

```

This provides standard summary measures of model fit, such as RMSE.

The purpose of this step is to assess whether the addition or removal of model
terms leads to a materially different fit.

In practice, this comparison is often used to support modelling choices before
moving to tariff interpretation.


## Step 2 — Coefficient inspection

Model validation is not limited to summary fit statistics. The coefficient
structure also needs to be reviewed.

```{r}

rating_table(mod1, mod2, model_data = df, exposure = "exposure") |>
  autoplot()

```

This is used to assess:

-	the relative size of fitted effects
-	the ordering of factor levels
-	the exposure behind each level
-	whether differences are plausible and stable

In pricing practice, this is a standard part of validation, because a model with
slightly better fit may still be less suitable if its coefficient structure is
difficult to interpret or unstable in low-exposure segments.


## Step 3 — Predictive stability

Single performance measures provide only a point estimate. In many pricing
contexts, it is also relevant to assess how stable that performance is under
small variations in the data.

```{r}

bootstrap_performance(mod1, df, n = 100, show_progress = FALSE) |>
  autoplot()

```

This evaluates predictive stability by repeatedly refitting the model on
bootstrap samples and storing the resulting RMSE values.

The output is used to assess:

-	the variability of model performance
-	whether the fitted model behaves consistently
-	whether the model is highly sensitive to changes in the underlying sample

This is particularly relevant when portfolios contain sparse segments or large
claim volatility.


## Step 4 — Dispersion checks

For Poisson models, it is standard practice to check whether the variance
assumption is broadly appropriate.

```{r}

check_overdispersion(mod1)

```

A dispersion ratio above 1 indicates that the observed variance exceeds the
variance implied by the Poisson model.

This does not automatically invalidate the model, but it does provide an
important diagnostic signal. In pricing practice, overdispersion may indicate:

-	omitted heterogeneity
-	model misspecification
-	clustering in the data
-	or unmodelled portfolio structure


## Step 5 — Residual diagnostics

Residual diagnostics provide an additional view of model adequacy.

```{r}

check_residuals(mod1, n_simulations = 600) |>
  autoplot()

```

This step is used to assess whether the residual behaviour is broadly consistent
with the fitted model assumptions.

In GLM settings, simulation-based residual diagnostics are often more useful
than classical residual plots, because they allow the fitted model to be
evaluated relative to its own implied distribution.

The purpose of this step is not to search for perfect residual behaviour, but to
identify material deviations that may be relevant for pricing.


## Step 6 — Portfolio-level structure

Validation is also performed at portfolio or model-point level.

```{r}

grid <- rating_grid(mod1)
head(grid)

```

`rating_grid()` aggregates the fitted model to observed model-point
combinations. This is useful when validation requires a more structured view of:

-	observed portfolio composition
-	combinations of rating factors
-	model-point level summaries
-	compact portfolio input for further review

This step is particularly relevant when moving from model validation to tariff
review or implementation support.


## Validation in context

In practice, model validation is rarely based on a single statistic.

A standard validation workflow combines:

-	comparative performance measures
-	coefficient inspection
-	predictive stability
-	residual and dispersion diagnostics
-	portfolio-level review

These steps serve different purposes:

-	performance measures assess fit
-	coefficient inspection assesses interpretability
-	bootstrap analysis assesses stability
-	diagnostics assess model adequacy
-	portfolio review assesses practical usability

Taken together, they provide a structured basis for deciding whether a model is
suitable for pricing use.


## Summary

A standard validation workflow in `insurancerating` is:

```{r, eval = FALSE}

model_performance(...)        # compare fitted models
rating_table(...) |> autoplot()   # inspect coefficient structure
bootstrap_performance(...)    # assess predictive stability
check_overdispersion(...)     # assess dispersion
check_residuals(...)          # inspect residual behaviour
rating_grid(...)              # review model-point structure

```

The purpose of validation is not only to assess model fit, but to determine
whether the fitted model is:

-	statistically adequate
-	stable under data variation
-	interpretable in tariff terms
-	suitable for practical pricing use


## Next steps

For the standard modelling workflow, see:

- [Getting started](getting-started.html)

For the refinement step after validation, see:

- [Refinement workflow](refinement-workflow.html).

For the conceptual background to exposure, risk premium, and tariff structure,
see:

- [Pricing principles](pricing-principles.html)
