---
title: "Refinement building blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Refinement building blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

In many pricing analyses, model estimation is followed by a translation step.

A fitted GLM may capture the structure of the portfolio well, while some fitted
effects still need to be reviewed before they are used in a tariff.

Common reasons include:

-	irregular local variation
-	lack of monotonicity
-	externally imposed tariff structures
-	expert judgement not directly represented in the model
-	implementation constraints in policy administration systems

For this reason, actuarial pricing work often distinguishes between:

1.	model estimation
2.	tariff refinement
3.	final refit of the pricing structure

`insurancerating` provides a staged refinement interface:

1.	fit an unrestricted model
2.	initialise a refinement object with `prepare_refinement()`
3.	add one or more refinement steps
4.	inspect these steps before refit
5.	call `refit()` to obtain the final fitted model

This separation records which adjustments are proposed before they are included
in the fitted model. It also distinguishes estimated GLM effects from explicit
tariff assumptions.


## When refinement can help

Refinement can be considered when the estimated model captures the main risk
structure, but selected coefficient patterns require additional structure
before tariff implementation.

Typical use cases include:

-	smoothing a rating factor derived from a continuous variable
-	imposing monotonicity
-	restricting coefficients to a predefined relativity structure
-	introducing expert-based relativities within existing model levels
-	simplifying the final tariff for practical implementation

In many workflows, refinement is applied to the model that represents the final
pricing signal, such as a premium or risk-premium model. In other cases, it may
also be useful for selected frequency or severity effects. The relevant question
is whether the adjusted coefficient pattern is intended to support the tariff
structure that will be reviewed or implemented.


## Example setup

The example below starts from one common premium modelling setup:

-	analyse a continuous variable with a GAM
-	convert it to tariff segments
-	fit frequency and severity models
-	combine both into a premium proxy
-	fit an unrestricted premium model
	

```{r, message = FALSE, warning = FALSE}

library(insurancerating)
library(dplyr)

age_policyholder_frequency <- risk_factor_gam(
  data = MTPL,
  claim_count = "nclaims",
  risk_factor = "age_policyholder",
  exposure = "exposure"
)

age_segments_freq <- derive_tariff_segments(
  age_policyholder_frequency,
  seed = 1
)

dat <- MTPL |>
  add_tariff_segments(age_segments_freq, name = "age_policyholder_freq_cat") |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(across(where(is.factor), ~ set_reference_level(., exposure)))

freq <- glm(
  nclaims ~ bm + age_policyholder_freq_cat,
  offset = log(exposure),
  family = poisson(),
  data = dat
)

sev <- glm(
  amount ~ zip,
  weights = nclaims,
  family = Gamma(link = "log"),
  data = dat |> filter(amount > 0)
)

premium_df <- dat |>
  add_prediction(freq, sev) |>
  mutate(premium = pred_nclaims_freq * pred_amount_sev)

burn_unrestricted <- glm(
  premium ~ zip + bm + age_policyholder_freq_cat,
  weights = exposure,
  family = Gamma(link = "log"),
  data = premium_df
)

```

Before refinement, inspect the unrestricted coefficient structure:

```{r}

rating_table(burn_unrestricted)

rating_table(burn_unrestricted) |>
  autoplot()

```

At this stage, the coefficients reflect the unrestricted model fit. This output
is often informative by itself. If the pattern is too irregular, too granular or
difficult to explain, a refinement step can be added explicitly.


## The refinement object

Refinement begins with:

```{r}

ref <- prepare_refinement(burn_unrestricted)
ref

```

A `rating_refinement` object stores:

-	the fitted base model
-	the underlying model data
-	the refinement steps added through the refinement interface

At this point, the model itself has not been refitted. The refinement object
represents a proposed tariff adjustment structure, not yet the final fitted
result.

This distinction is useful because refinement steps can be inspected before they
are incorporated into the final model.


## Smoothing 

### Purpose

Smoothing can be used when a rating factor derived from a continuous variable
contains local variation that is hard to justify in a tariff.

For example, a coefficient pattern such as:

-	age 30–34 lower
-	age 34–38 higher
-	age 38–42 lower again

may be compatible with the observed sample, but unstable across periods or
difficult to support actuarially. Smoothing replaces local variation with an
explicitly regularised coefficient pattern.


### Adding smoothing

```{r}

ref <- ref |>
  add_smoothing(
    model_variable = "age_policyholder_freq_cat",
    source_variable = "age_policyholder",
    breaks = c(seq(18, 93, 5), 95),
    weights = "exposure"
  )

```

The key arguments are:

-	`model_variable`: the grouped variable present in the GLM
-	`source_variable`: the original continuous portfolio variable
-	`breaks`: the preferred commercial cut points
-	`smoothing`: the smoothing specification; `"spline"` is the general-purpose
  default
-	`weights`: optional weighting, typically exposure

Use `"increasing"` or `"decreasing"` when the tariff effect is required to
move monotonically. Convex, concave and combined shape constraints are
available for cases where that additional assumption can be supported
actuarially. `"poly"` fits a global polynomial and uses `degree`; `"gam"` fits
a thin-plate smooth that can be used as an unconstrained comparison. For spline
methods, `k` limits the available curve flexibility but is not the fitted
effective number of degrees of freedom.

### Inspecting smoothing before refit

```{r}

print(ref)
autoplot(
  ref,
  variable = "age_policyholder_freq_cat",
  x_max = 90,
  y_max = 1.5
)

```

This plot belongs to the **pre-refit stage**. It shows:

-	the original fitted coefficients
-	the proposed smoothed structure

The purpose is to inspect the refinement step itself, before it is incorporated
into the final fitted model.

### Choosing a smoothing method

Typical smoothing choices are:

-	`"spline"`: unconstrained penalised cubic regression spline; the
  general-purpose default
-	`"poly"`: global polynomial controlled by `degree`
-	`"gam"`: thin-plate regression spline fitted with `mgcv`, mainly useful as
  an unconstrained comparison
-	`"increasing"`: monotone increasing
-	`"decreasing"`: monotone decreasing

Monotonicity concerns the direction of the tariff effect. Convexity and
concavity concern how its slope changes. A convex effect has an increasing
slope and may be U-shaped when its direction is not constrained; a concave
effect has a decreasing slope and may be inverted U-shaped. For an increasing
effect, convexity implies acceleration and concavity implies flattening. For a
decreasing effect, convexity implies flattening and concavity implies a steeper
decline.

The readable advanced options are `"convex"`, `"concave"`,
`"increasing_convex"`, `"increasing_concave"`, `"decreasing_convex"` and
`"decreasing_concave"`. They should be used only when the assumed curvature
has an actuarial or economic basis in addition to any directional assumption.
The former short codes `"mpi"`, `"mpd"`, `"cx"`, `"cv"`, `"micx"`, `"micv"`,
`"mdcx"` and `"mdcv"` remain accepted for compatibility.

For example:

-	age may justify a flexible smooth
-	insured value or power may require a monotonic relationship
-	low-exposure tails may benefit from exposure weighting


## Restrictions

### Purpose

Restrictions can be used when coefficients need to follow a predefined
structure.

Typical examples include:

-	bonus-malus systems
-	governance-approved relativities
-	externally mandated tariff structures
-	implementation constraints in policy systems

Restrictions differ from smoothing:

-	smoothing reshapes the fitted pattern
-	restriction imposes user-defined coefficients


### Adding restrictions



```{r}

zip_df <- data.frame(
  zip = c(0, 1, 2, 3),
  zip_adj = c(0.8, 0.9, 1.0, 1.2)
)

ref <- ref |>
  add_restriction(restrictions = zip_df)

```

The restriction table must contain exactly two columns:

-	the original factor levels
-	the adjusted coefficients


### Inspecting restrictions before refit

```{r}

autoplot(ref, variable = "zip")

```

This shows the proposed restricted structure relative to the original fitted
model.


## Expert-based relativities

### Purpose

In some cases, the fitted model uses a broad factor level, while portfolio or
business knowledge suggests that more granular differentiation may be useful.

For example, a model may estimate one coefficient for "construction", while
pricing practice distinguishes between:

-	residential construction
-	commercial construction
-	civil engineering

This can be relevant when subgroup exposure is too limited to estimate stable
coefficients directly.

### Adding relativities

```{r, eval = FALSE}

relativities_activity <- relativities(
  split_level(
    "construction",
    c("residential_construction", "commercial_construction"),
    c(1.00, 1.15)
  ),
  split_level(
    "services",
    c("professional_services", "personal_services"),
    c(0.95, 1.05)
  )
)

ref <- ref |>
  add_relativities(
    model_variable = "business_activity",
    split_variable = "business_activity_split",
    output_variable = "business_activity_tariff_segment",
    relativities = relativities_activity,
    exposure = "exposure",
    normalize = TRUE
  )

```

If `normalize = TRUE`, the relativities are scaled so that their
exposure-weighted average remains equal to 1 within the original level.

`output_variable` names the resulting hybrid tariff factor. In this example,
the explicitly split construction and services levels are represented by their
detailed levels, while unsplit levels retain their original
`business_activity` value. A name ending in `_tariff_segment` can make this
role explicit in later model and reporting steps.

This preserves the original model signal while introducing finer structure.
When `model_variable` has already been restricted with `add_restriction()`, the
restricted coefficients are used automatically as the basis for these
relativities. Refinement order is therefore part of the specification.


## Refit

### Why refit is required

Refinement steps alter part of the model structure. Once these changes are
applied, the remaining coefficients may also adjust.

For that reason, the sequence does not end with `add_smoothing()` or
`add_restriction()`. The final step is:

```{r}

summary(ref)

burn_refined <- refit(ref)

```

`summary(ref)` describes the base model and the ordered refinement specification
before a new GLM is fitted. This provides a direct review of the assumptions
that will be applied. `refit()` then fits the model while incorporating those
documented steps.


### Inspecting the final fitted result

After refit, use `rating_table()`:

```{r}

rating_table(burn_refined)

```

At this point, the output no longer represents a proposed refinement plan. It
represents the fitted coefficient structure after refinement.

The distinction is:

-	before `refit()` --> inspect the refinement plan
-	after `refit()` --> inspect the fitted tariff structure

If smoothing, restrictions, and relativities have been applied, they are now
embedded in the fitted model output.


### Auditing the portfolio effect

Coefficient changes cannot be interpreted independently of the intercept and
the other model terms. A more direct audit compares fitted values from the
unrestricted and refined models on the same observed portfolio combinations.

```{r}

refinement_audit <- audit_refinement(
  burn_refined,
  exposure = "exposure",
  metric = "risk_premium"
)

summary(refinement_audit)

```

The portfolio result shows the exposure-weighted risk premium before and after
refinement. `as.data.frame(refinement_audit)` gives the corresponding comparison
for every final risk factor and level. These level results reflect the combined
prediction of the full model for the observed portfolio mix; they are not
isolated coefficient differences.

The audit also records the package version, preparation time, refit time,
formulas and ordered refinement steps. For a frequency or severity model, use a
matching metric name such as `"frequency"` or `"average_severity"`; a complete
risk-premium interpretation requires a direct risk-premium model or a combined
frequency and severity calculation.


### Visualising the final structure

```{r}

rating_table(burn_refined) |>
  autoplot()

```



## Model data and rating grids 

After refit, model structure can be extracted with `extract_model_data()`:

```{r}

md <- extract_model_data(burn_refined)
head(md)

```

A model point represents a unique observed combination of the risk-factor
levels used by the fitted model. `rating_grid()` groups portfolio records with
the same combination and attaches the corresponding refined model effects. It
returns combinations observed in the portfolio; it does not construct every
theoretically possible combination of factor levels.

Observed model-point combinations can be obtained with `rating_grid()`:

```{r}

grid <- rating_grid(burn_refined)
head(grid)

```

Each row can represent several portfolio records. `count` gives the number of
records with that combination and `exposure` gives their aggregated exposure.
The remaining refinement columns contain the restrictions, smoothed effects or
derived relativities that apply to the model point.

When a fitted GLM is supplied, `rating_grid()` retrieves the exposure column
used as a model weight or offset from the model metadata and aggregates it
automatically. Supply `exposure` explicitly only when that column is not used by
the model or when a plain data frame is supplied.

This is typically used for:

-	tariff review
-	portfolio summaries
-	compact input for tariff calculations or predictions
-	implementation support

Before the grid is used for prediction, the analyst should verify that all
required model variables are present and decide how combinations not observed
in the estimation portfolio will be handled.
	
	
## Complete example

One possible refinement sequence is:

```{r}

zip_df <- data.frame(
  zip = c(0, 1, 2, 3),
  zip_adj = c(0.8, 0.9, 1.0, 1.2)
)

burn_refined <- prepare_refinement(burn_unrestricted) |>
  add_smoothing(
    model_variable = "age_policyholder_freq_cat",
    source_variable = "age_policyholder",
    breaks = c(seq(18, 93, 5), 95),
    weights = "exposure"
  ) |>
  add_restriction(zip_df) |>
  refit()

rating_table(burn_refined)

rating_table(burn_refined) |>
  autoplot()

```


## Legacy interface

Legacy entry points remain available:

```{r, eval = FALSE}

burn_refined_old <- burn_unrestricted |>
  smooth_coef(
    x_cut = "age_policyholder_freq_man",
    x_org = "age_policyholder",
    breaks = c(seq(18, 93, 5), 95)
  ) |>
  restrict_coef(zip_df) |>
  refit_glm()

```

These are primarily maintained for backward compatibility.

For new code, the recommended interface is:

```{r, eval = FALSE}

prepare_refinement() |> add_*() |> refit()

```

This keeps the sequence of tariff adjustments explicit.



## Summary

The refinement interface helps separate:

-	model estimation
-	tariff adjustments
-	final fitted output

The refinement specification records which parts of the final coefficient
structure originate from the fitted model and which parts result from
smoothing, restrictions or expert-based sublevel relativities. These choices
should be assessed using exposure, claim volume, stability and the intended
tariff application.


## Next steps

For the underlying pricing concepts, see:

- [Pricing workflow building blocks](pricing-workflow-building-blocks.html)

For an example sequence from portfolio analysis to fitted tariff, see:

- [Getting started](getting-started.html)
