---
title: "Refinement building blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Refinement building blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Refinement is the explicit translation from statistically estimated model
effects to the structure that will be reviewed, justified and implemented as a
tariff. It can be motivated by stability, credibility, monotonicity,
smoothness, sparse experience or an explicit implementation constraint.

Refinement is not arbitrary editing of inconvenient coefficients. Each
adjustment should have an actuarial or operational rationale and should be
reviewed against exposure, claim volume, observed experience and model
diagnostics. Recording these decisions as ordered steps makes them
reproducible and easier to review.

This vignette starts from a fitted unrestricted GLM and focuses on what happens
next. The construction of frequency, severity and technical risk-premium
models is covered in [Getting Started](getting-started.html). The wider role of
refinement within the package is mapped in
[Pricing workflow and package building blocks](pricing-workflow-building-blocks.html).

The refinement architecture is:

| Stage | Meaning |
|---|---|
| Unrestricted GLM | Statistical starting point |
| `prepare_refinement()` | Create an editable specification around that model |
| `add_smoothing()`, `add_restriction()`, `add_shrinkage()`, `add_relativities()` | Record proposed actuarial adjustments |
| `summary()` and `autoplot()` | Inspect the proposal before fitting |
| `refit()` | Reconstruct and fit the GLM under the stored specification |
| `rating_table()` and `audit_refinement()` | Inspect the fitted tariff and its portfolio effect |

## A compact unrestricted model

The unrestricted model is only the starting material for this vignette. The
following setup creates grouped age and bonus-malus factors and fits a Poisson
frequency GLM. The response is claim count and `log(exposure)` is the offset.

```{r, message = FALSE, warning = FALSE}
library(insurancerating)

portfolio <- as.data.frame(MTPL)

age_breaks <- c(18, 25, 32, 39, 51, 58, 65, 84, 95)
portfolio$age_band <- cut(
  portfolio$age_policyholder,
  breaks = age_breaks,
  include.lowest = TRUE
)
portfolio$bm_group <- cut(
  portfolio$bm,
  breaks = c(0, 4, 8, Inf),
  labels = c("Low", "Medium", "High")
)
portfolio$bm_detail <- factor(as.character(portfolio$bm))
portfolio$zip <- factor(portfolio$zip)

unrestricted <- glm(
  nclaims ~ age_band + zip + bm_group + offset(log(exposure)),
  family = poisson(),
  data = portfolio
)
```

The initial tariff effects can be inspected before any actuarial adjustment:

```{r}
head(rating_table(unrestricted, exposure = FALSE))
```

These are the conditional effects estimated by the unrestricted GLM. An
irregular effect is not by itself a reason to refine it: the analyst should
first consider data quality, exposure definition, model specification,
interactions and factor construction.

## Preparing the refinement

```{r}
refinement <- prepare_refinement(
  unrestricted,
  data = portfolio
)

refinement
```

`prepare_refinement()` creates a `rating_refinement` object containing the
unrestricted model, the corresponding retained model data and an initially
empty ordered step specification. It does not refit the GLM and does not change
its coefficients or fitted values.

This distinction is central to the API:

- `unrestricted` is a fitted statistical model;
- `refinement` is an editable proposal for the tariff structure;
- the model returned later by `refit()` is the fitted outcome under that
  proposal.

Keep the refinement object when iterating. A model returned by `refit()` is a
fitted result, not an editable specification. Calling `prepare_refinement()` on
that fitted result deliberately starts a new workflow and does not reconstruct
the earlier steps.

## Smoothing an ordered effect

Raw level effects for a grouped continuous variable may contain local movement
that is weakly supported or unstable over time. Smoothing replaces that local
pattern with an explicitly structured curve. It is most relevant when a gradual
underlying relationship is plausible and neighbouring tariff levels should be
interpreted coherently.

```{r}
refinement <- refinement |>
  add_smoothing(
    model_variable = "age_band",
    source_variable = "age_policyholder",
    breaks = age_breaks,
    smoothing = "spline",
    k = 5,
    weights = "exposure"
  )
```

The arguments distinguish two variables:

- `model_variable` is the grouped factor included in the unrestricted GLM;
- `source_variable` is the underlying numeric portfolio variable used to
  estimate the smooth relationship.

`breaks` define the intervals in the resulting tariff factor. Exposure weights
give levels with more portfolio support more influence. With a spline method,
`k` controls the available curve flexibility; it is a basis dimension rather
than a fixed number of fitted degrees of freedom.

The default unconstrained spline is suitable when the shape should remain
data-led. Increasing or decreasing shape constraints are available when a
directional assumption has a defensible actuarial basis. The function reference
for `add_smoothing()` describes the full set of methods and their curvature
interpretation.

### Inspecting the proposal

```{r}
summary(refinement)

autoplot(
  refinement,
  variable = "age_band",
  x_max = 90
)
```

This is a **pre-refit** plot. It compares the original fitted effect with the
proposed smooth structure. The GLM has not yet been estimated under that
structure.

If the curve needs adjustment, retain the specification and use
`edit_smoothing()` before refitting. This preserves the rest of the ordered
workflow rather than starting again from the fitted result.

### Refining a selected part of the smoothing

`add_smoothing()` defines the initial curve and its structural assumptions.
`edit_smoothing()` can subsequently refine a selected interval in two distinct
ways. These edits are explicit tariff assumptions layered on the initial
smoothing; they are not additional statistical observations.

Each call to `edit_smoothing()` is stored as a separate refinement step. If the
same smoothing is edited more than once, the edits are applied cumulatively in
their recorded order. The `step` argument in `autoplot()` can therefore be used
to inspect the curve after a particular edit without introducing a separate
revision argument.

#### Explicitly redirecting the curve

Use explicit values when the intended targets are known:

```{r}
explicit_age_refinement <- refinement |>
  edit_smoothing(
    model_variable = "age_band",
    from = 32,
    to = 65,
    from_value = 0.95,
    to_value = 1.10,
    control_positions = 51,
    control_values = 1.02
  )

autoplot(explicit_age_refinement, variable = "age_band", x_max = 90)
```

The stored edit redirects the selected interval through the supplied target
values. This is appropriate when those values have a documented actuarial or
implementation basis.

#### Applying a relative local adjustment

Use `adjustment` when the current shape is broadly appropriate but one region
should be somewhat higher or lower:

```{r}
relative_age_refinement <- refinement |>
  edit_smoothing(
    model_variable = "age_band",
    from = 32,
    to = 65,
    adjustment = 1.05
  )

autoplot(
  relative_age_refinement,
  variable = "age_band",
  x_max = 90,
  show_initial_smoothing = TRUE
)
```

Here, `adjustment = 1.05` raises the middle of the selected region by up to 5%
relative to the existing smoothing. The multiplier starts at 1 at `from`, moves
towards 1.05 over the available smoothing points, and returns to 1 at `to`.
The unchanged curve is therefore retained outside the interval and no jump is
introduced at either boundary.

With `show_initial_smoothing = TRUE`, the plot includes both the curve directly
after `add_smoothing()` and the cumulative curve at the selected step. This
comparison changes only the plot; it does not alter the stored refinement or
the subsequent refit.

For example, two local adjustments create two successive edit steps:

```{r}
cumulative_age_refinement <- relative_age_refinement |>
  edit_smoothing(
    model_variable = "age_band",
    from = 50,
    adjustment = 1.02,
    transition = "linear"
  )

# Step 3 is the cumulative result of the initial smoothing and both edits.
autoplot(
  cumulative_age_refinement,
  step = 3,
  x_max = 90,
  show_initial_smoothing = TRUE
)
```

Selecting `step = 2` would show the curve after the first edit only. The
initial comparison line remains the result of the original `add_smoothing()`
step in both plots.

For a tail adjustment, only one boundary is needed. With only `from`, the
adjustment runs from that value to the end of the available smoothing range.
With only `to`, it runs from the beginning of the range to that value. The
transition remains attached to the original curve at the supplied boundary:

```{r, eval = FALSE}
# Refine the upper tail
edit_smoothing(
  refinement,
  model_variable = "age_band",
  from = 50,
  adjustment = 1.05
)

# Refine the lower tail
edit_smoothing(
  refinement,
  model_variable = "age_band",
  to = 35,
  adjustment = 0.95
)
```

With `transition = NULL`, the default, the transition inherits the smoothing
specification of the step being edited. For example, an original
`smoothing = "increasing_concave"` uses the same structural shape logic while
adapting the entry and exit to their opposite directions. The resulting curve
is checked against the inherited monotonicity and curvature where applicable.

Two explicit alternatives are available:

```{r, eval = FALSE}
# Continuous straight transitions
edit_smoothing(
  refinement,
  model_variable = "age_band",
  from = 32,
  to = 65,
  adjustment = 1.05,
  transition = "linear"
)

# Immediate changes at both boundaries
edit_smoothing(
  refinement,
  model_variable = "age_band",
  from = 32,
  to = 65,
  adjustment = 1.05,
  transition = "step"
)
```

`"linear"` remains continuous. `"step"` deliberately permits discontinuities
and should therefore be selected only when an immediate tariff change is
intended. Relative adjustments and explicit target values cannot be combined
in one `edit_smoothing()` call because they express different instructions.
They can be recorded in separate consecutive edit steps when that sequence has
a clear actuarial interpretation.

The remainder of this vignette uses `relative_age_refinement` as the current
proposal:

```{r}
refinement <- relative_age_refinement
```

## Restricting selected levels

`add_restriction()` fixes tariff levels at supplied relativities. A restriction
may represent an implementation rule, a supported external assumption or a
deliberate response to an unstable local estimate. It differs from smoothing:
smoothing estimates a structured pattern, whereas a restriction explicitly
prescribes selected values.

```{r}
zip_restrictions <- data.frame(
  zip = c("0", "3"),
  zip_restricted = c(0.95, 1.10)
)

refinement <- refinement |>
  add_restriction(zip_restrictions)
```

Only ZIP levels 0 and 3 are supplied here. The other observed ZIP levels are
fixed at their current effective relativities. Consequently, a partial
restriction changes the selected values but still produces a complete fixed
structure for that risk factor.

Repeated calls for the same restricted variable update matching levels and
retain earlier restrictions for levels not supplied again. New levels may be
added when they represent explicit tariff assumptions; new risk factors require
the corresponding portfolio column and an explicit opt-in. These behaviours
are documented in the `add_restriction()` reference.

A new fixed tariff factor can also replace an existing standalone model term.
For example, `add_restriction(..., replaces = "postal_area")` records that the
new classification substitutes for `postal_area` during `refit()`, rather than
adding a second multiplicative effect. The replacement is shown in the
refinement summary and audit. Interactions and transformed terms must be revised
explicitly in the model specification because removing them is not an
unambiguous level restriction.

```{r}
autoplot(refinement, variable = "zip")
```

Again, the plot shows the proposed restriction before `refit()`.

## Shrinking a categorical effect

`add_shrinkage()` reduces the differences between the current relativities of
one categorical risk factor. It is useful when the direction and ordering of an
estimated effect are considered informative, but the spread between levels is
larger than is supported by the available experience or the intended tariff.
Unlike a restriction, shrinkage does not prescribe individual level values.

```{r}
refinement <- refinement |>
  add_shrinkage(
    model_variable = "bm_group",
    credibility = 0.9,
    weights = "exposure"
  )
```

`credibility` is the weight assigned to the current logarithmic effect. A value
of 0.9 retains 90 percent of each level's deviation from the common centre; a
value of 1 leaves the effect unchanged, while a value of 0 removes differences
between levels. This is a user-selected refinement parameter, not an
automatically estimated Buhlmann or Buhlmann-Straub credibility factor.

The common centre is determined using the selected weighting basis. Exposure
weights are appropriate here because `bm_group` is part of a frequency model.
For a severity model, claim count may be a more relevant basis. Use
`weights = "equal"` when every factor level should receive equal weight rather
than representing the observed portfolio mix.

After shrinkage, the relativities are normalised so their exposure-weighted
arithmetic mean remains equal to its value before shrinkage. The operation
therefore reduces tariff differentiation without intentionally changing the
weighted level of this risk factor. The later refit can still recalibrate the
intercept or other free model effects.

```{r}
autoplot(refinement, variable = "bm_group")
```

This remains a pre-refit comparison: it shows the current estimated effect and
the proposed shrunken structure stored in the refinement specification.

## Adding differentiation within model levels

`add_relativities()` addresses a different problem. An unrestricted GLM may use
a broad factor because its detailed levels are individually too sparse for
stable direct estimation. The refinement can retain the broad parent effect
while introducing documented differentiation between selected sublevels.

The example splits the broad bonus-malus groups `Low` and `Medium` into their
observed detailed values:

```{r}
bm_relativities <- relativities(
  split_level(
    "Low",
    new_levels = c("1", "2", "3", "4"),
    relativities = c(0.95, 0.98, 1.02, 1.05)
  ),
  split_level(
    "Medium",
    new_levels = c("5", "6", "7", "8"),
    relativities = c(0.96, 0.99, 1.02, 1.05)
  )
)

refinement <- refinement |>
  add_relativities(
    model_variable = "bm_group",
    split_variable = "bm_detail",
    relativities = bm_relativities,
    exposure = "exposure",
    normalize = TRUE,
    output_variable = "bm_tariff_segment"
  )
```

`model_variable` supplies the parent GLM effect. `split_variable` identifies the
detailed portfolio levels within each parent. `output_variable` names the
resulting hybrid tariff factor; unsplit parent levels retain their existing
model effect.

With `normalize = TRUE`, the sublevel relativities are normalised within each
parent so their exposure-weighted average equals one. The split therefore
redistributes the parent effect without changing its exposure-weighted level.
Because shrinkage was added first, these sublevel effects use the shrunken
`bm_group` relativities as their parent values. With `normalize = FALSE`, the
supplied relativities are applied directly.

This operation is not equivalent to restriction or smoothing:

- smoothing regularises an ordered effect already represented by the model;
- restriction fixes selected tariff values;
- shrinkage reduces differences between categorical levels while retaining
  their ordering;
- additional relativities introduce finer differentiation inside a broader
  model level.

Step order matters. A restriction or shrinkage step added before
`add_relativities()` changes the parent coefficient used as the basis for the
split. A later restriction can instead adjust selected levels of the derived
`output_variable`.

## Combining and reviewing refinements

The four operations now form one ordered specification:

```{r}
summary(refinement)
```

The summary records the base formula, package version, data size and each
refinement in evaluation order. It describes what will be applied; it does not
report a fitted refined model because `refit()` has not yet been called.

The variable-specific `autoplot()` calls above evaluate the stored steps in
their recorded order. They therefore inspect the intended tariff after any
preceding adjustments without silently changing the GLM. The `step` argument
can be used when an earlier stage of a longer specification needs to be
reviewed separately.

`add_rebasing()` can be inserted when required to change which resulting level
is displayed as one without changing ratios between levels. Rebasing changes
the representation of an effect, whereas shrinkage changes its degree of
differentiation.

## Refitting the model

```{r}
refined_model <- refit(
  refinement,
  intercept_only = TRUE
)
```

`refit()` applies the stored steps in order, constructs the required tariff
variables and offsets, updates the formula and calls `glm()` with the original
model family. It does more than copy proposed relativities into an existing
coefficient vector.

The treatment of the remaining model effects depends on `intercept_only`:

- with `intercept_only = TRUE`, unaffected existing effects are fixed as
  offsets and only the intercept is estimated. Their relative differences are
  preserved while the overall level is recalibrated;
- with `intercept_only = FALSE`, remaining free terms are estimated again.
  Their coefficients may change as the GLM finds a new joint optimum conditional
  on the fixed refinement steps.

An intercept-only refit is often appropriate for a controlled adjustment to an
accepted tariff structure. Re-estimating the remaining free effects is more
appropriate when the refinement is part of substantive model development and
dependence between factors should be reconsidered.

The prescribed smoothing, restrictions and sublevel relativities remain
explicit assumptions. Refitting does not turn them into unrestricted statistical
estimates.

## Inspecting the fitted result

After `refit()`, the result is a fitted GLM and can be reviewed with the normal
interpretation tools:

```{r}
head(rating_table(refined_model, exposure = FALSE))
```

This is the **post-refit** result. It is distinct from the preview shown by
`autoplot(refinement)`:

- before `refit()`: proposed tariff adjustments;
- after `refit()`: fitted model under those adjustments.

Printing `refined_model` also reports the original and refitted formulas, the
model family, refit mode and stored refinement steps before showing the regular
GLM output.

## Auditing the portfolio effect

Individual coefficient changes are difficult to interpret independently of
the intercept, other model terms and portfolio mix. `audit_refinement()`
therefore compares predictions from the unrestricted and refined models on the
same observed portfolio combinations.

```{r}
refinement_audit <- audit_refinement(
  refined_model,
  exposure = "exposure",
  metric = "frequency"
)

summary(refinement_audit)
```

The audit reports the portfolio-level frequency before and after refinement
and the corresponding changes by final risk-factor level. These are predictions
from the complete model for the observed portfolio mix, not isolated
coefficient differences.

The refinement and audit objects record package version, timestamps, formulas
and ordered steps. This supports reproducibility, peer review and documentation
of actuarial judgement. Organisational approval, source-data versioning and
formal governance remain outside the package and should be handled by the
user's normal processes.

## When to revisit the model

Refinement is not a substitute for correcting a poor model. If an implausible
effect is caused by incorrect data, exposure errors, missing interactions, an
inappropriate response definition or poor factor construction, the model or
data preparation should be revisited first. Refinement is most defensible when
the statistical model captures the main risk structure and the remaining
adjustment has a clear tariff rationale.

## Iterating without losing the specification

Retain both the editable specification and the fitted result:

```{r, eval = FALSE}
refined_model <- refit(refinement)

refinement <- refinement |>
  edit_smoothing(
    model_variable = "age_band",
    from = 32,
    to = 65,
    adjustment = 1.03,
    transition = "linear"
  )

updated_model <- refit(refinement)
```

The update replaces the earlier local adjustment on this smoothing step. It is
always calculated relative to the initial smoothing and is therefore not
multiplied cumulatively by the previous value.

Calling `add_smoothing()`, `add_restriction()` or another refinement function
directly on `refined_model` is deliberately not supported. Further adjustments
belong on the retained `rating_refinement` object so their ordering and origin
remain visible.

## Summary

The specialist workflow is:

1. start from an unrestricted fitted model;
2. create one persistent specification with `prepare_refinement()`;
3. add smoothing, restrictions, shrinkage or sublevel relativities with an
   explicit rationale;
4. inspect the proposal with `summary()` and `autoplot()`;
5. use `refit()` to fit the model under that specification;
6. inspect the final tariff with `rating_table()` and its portfolio effect with
   `audit_refinement()`.

This separates estimated effects, proposed actuarial adjustments and final
fitted output without treating refinement as an automatic approval of the
result.

## Where to go next

- [Getting Started](getting-started.html) constructs the initial pricing model
  used as the starting point for refinement.
- [Pricing workflow and package building blocks](pricing-workflow-building-blocks.html)
  places refinement within the wider package architecture.
- [Model validation](model-validation.html) develops the diagnostics used to
  assess unrestricted and refined models.
